Data exploration through dot-driven development\footnote{This work was supported by Google Digital News Initiative.}

### 2017-01-11 pivot/paper.tex
Case study

\begin{figure}
\begin{center}
\includegraphics[scale=0.35,trim=0mm 0mm 0mm 0mm,clip]{filter.png} % left bottom right top
\end{center}
\caption{Auto-completion offering the available values of the athlete name column}
\label{fig:dot-driven}
\end{figure}

@@

\section{Implementation}
\label{sec:impl}

Rio case study

\begin{equation*}
\begin{array}{l}
\ident{olympics}\\
\quad.\qident{filter data}.\qident{Games is}.\qident{Rio (2016)}.\ident{then}\\
\quad.\qident{group data}.\qident{by Athlete}.\qident{sum Gold}.\ident{then}\\
\quad.\qident{sort data}.\qident{by Gold descending}.\ident{then}\\
\quad.\qident{paging}.\ident{take}(8)\\
\quad.\qident{get series}.\qident{with key Athlete}.\qident{and value Gold}
\end{array}
\end{equation*}

~
\newpage

>>>>

\begin{figure}
\begin{center}
\includegraphics[scale=0.35,trim=0mm 0mm 0mm 0mm,clip]{images/filter.png} % left bottom right top
\end{center}
\caption{Auto-completion offering the available values of the athlete name column}
\label{fig:dot-driven}
\end{figure}

@@

\begin{figure}[b]
\begin{center}
\includegraphics[scale=0.30,trim=0mm 0mm 0mm 0mm,clip]{images/table.png} % left bottom right top
\end{center}
\caption{Athletes by the number of medals over the enitre history of Olympic games}
\label{fig:case-table}
\end{figure}

\section{Case study: Visualizing Olympic medalists}
\label{sec:impl}

We used The Gamma script with the pivot type provider to build an interactive web site 
(\url{rio2016.thegamma.net}) that visualizes a number of facts about Olympic medalists 
using the data set discussed in Appendix~\ref{app:olympics-csv} and used throughout this paper.
The web site lets the readers view and modify the source code and we also developed a number of
tools that make working with the source code easier, going beyond the basic auto-completion tooling
to enable dot-driven development as discussed in Section~\ref{sec:analysis-auto}. In this section,
we reivew our experience and outline some of the additional tools (available at 
\url{github.com/the-gamma}).

\subparagraph{Building tables and charts.} 
Building rich visualizations with the simplicity of the pivot type provider was left for future
work -- we provided functions for building basic kinds of visualizations (table, column chart, 
pie chart and timeline) and we extended the host language with more advanced features that can
be used to customize the displays. Figure~\ref{fig:case-table} shows a sample table, listing
top athletes over the entire history of Olympic games.

The data transformation used to construct the table uses the operations discussed in this paper
together with paging functionality and \qident{get the data} which returns the entire data set
of type \ident{Query}, as opposed to extracting a series with keys and values:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{data}~=~\ident{olympics}\\
\quad .\qident{group data}.\qident{by Athlete}\\
\qquad .\qident{sum Gold}.\qident{sum Silver}.\qident{sum Bronze}.\qident{concat Team}.\ident{then}\\
\quad .\qident{sort data}.\qident{by Gold descending}\\
\qquad .\qident{and by Silver descending}.\qident{and by Bronze descending}.\ident{then}\\
\quad .\ident{paging}.\ident{take}(10).\qident{get the data}\\[0.5em]
\ident{table.create}(\ident{data})
\end{array}
\end{equation*}

\noindent
The $\ident{table.create}$ operation on the last line generates a table based on the columns 
available in the data set. We omit the additional customization which specifies that medals should
be rendered as images. For most visualizations we built, the pivot type provider was expressive
enough to capture the core logic of the operation, but further joining of data was sometimes
needed. Possible extensions that would allow capturing those are discussed in 
Section~\ref{sec:further}.

\begin{figure}[t]
\begin{center}
\includegraphics[scale=0.32,trim=0mm 0mm 0mm 0mm,clip]{images/options.png} % left bottom right top
\end{center}
\caption{User interface with automatically provided grouping and sorting options}
\label{fig:case-opts}
\end{figure}

\subparagraph{Generating interactive user interfaces.}
Although the pivot type provider simplifies code needed for data exploration, not every reader will 
be able to write or modify it. The simplicity of the host language makes it possible to easily 
generate user interface that allows changing of some of the parameters of the program. 
Figure~\ref{fig:case-opts} shows an example for the above code snippet that we implemented as 
part of the visualization.

The user interface lets the user choose aggregations to be caluclated over a group and select 
columns used for sorting. It is generated automatically by looking for specific pattern in the 
chain of member accesses -- we annotate members with annotations denoting whether a member is 
start of a list, list item or end of a list. The editor then looks for parts of the chain of the
form $\qident{list start}.\qident{list item 1}.\qident{list item 2}.\qident{list end}$ and generates
a control that lets the user remove or add list items. An item cannot be removed if the operation
caused the code to break (e.g.~when it adds a member that is needed later) and items to be added
are chosen using available members (as in the standard auto-complete). The header
shown in Figure~\ref{fig:case-opts} are provided as additional annotations attached to 
\qident{list start}.

\begin{figure}[t]
\begin{center}
\includegraphics[scale=0.31,trim=0mm 0mm 0mm 0mm,clip]{images/pivot.png} % left bottom right top
\end{center}
\caption{Auto-completion offering the available values of the athlete name column}
\label{fig:case-ed}
\end{figure}

\subparagraph{Spreadsheet-inspired live editor.}
The third editor extension that we developed for the pivot type provider aims to bridge the gap 
between code and user interface. This is done through a direct manipulation interface [X] inspired
by spreadsheet applications. When exploring data in a spreadsheet, the user can always see the 
data they work with and the results of an action will be immediately visible. For queries written
using the pivot type provider, the results can also be calculated immediately using the sample 
data set provided when using the refined version of the type provider as discussed in 
Section~\ref{sec:pivot-filter}.

The Figure~\ref{fig:case-ed} shows the sample expression (discussed above) in the live 
editor\footnote{The live editor can be tested live as part of the documentation for the 
JavaScript package at \url{thegamma.net}}.
Note that the selected part of code is the \qident{by Gold descending} identifier and so the 
preview shows data available at that point of the query evaluation. Athletes with largest number
of gold medals appear first, but silver or bronze medals are not used as secondary sorting keys yet.
As the user moves through the code, or writes the code, the live preview is updated accordingly.
Finally, the editor also makes it possible to modify the code through the user interface -- the
``x'' buttons can be used to remove sort keys or transformations and ``+'' buttons (on the right)
can be used to add more transformations or parameters.

Unlike the user interface for specifying lists, the live editor has been implemented specifically
for the pivot type provider, but it still heavily relies on the simple structure provided by the 
fact that entire transformation can be written as a single chain of member accesses. In particular,
we identify individual transformations (\qident{group by}, \qident{sort by}, etc.) and then 
generate different user interface for specifying parameters. For sorting, as shown in 
Figure~\ref{fig:case-ed}, the second level lets the user add or remove members. For grouping or
paging, the second level looks different and lets the user choose the key and the number of 
elements to take, respectively.

\section{Discussion and related work}

a

b

c

\subsection{Further work}
\label{sec:further}

\subsection{Related work}

visual query systems

https://people.csail.mit.edu/ebakke/research/sieuferd_sigmod2016.pdf

\subsection{Conclusions}
